---
title: "Practical Machine Learning Assignment"
author: "Chris Emerson"
date: "Saturday, October 24, 2015"
output: html_document
---

## Objective

To use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 
Report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

Details about data set and submission requirements can be found in the appendices

## Executive Summary 

Accuracy for Random Forest model was estimated at 0.9417038 making the expected out-of-sample error is estimated at 1-.9417 (.0538) Based on the cross-validation set, the  Accuracy is 0.9565 with 95% CI : (0.951, 0.9616) making the out of sample error 0.0435. Our Test data set comprises 20 cases. Based on these results, we can expect that very few, or none, of the test samples will be missclassified.

If time allows, I would like to understand why my attempts to use other methods generated errors, it would be nice to make a comparison.  But each iteration using a laptop has a severe time cost penalty. 

##Data Cleanup Exploratory Analyses
```{r getData,message=FALSE, warning=FALSE, cache=TRUE}
file_url_train <-  "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
file_url_test <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
file_name_train <- "./data/original/pml-training.csv"
file_name_test <-  "./data/original/pml-testing.csv"
if ( !file.exists(file_name_train) ) {
  dir.create("data/original")
  download.file(file_url_train, file_name_train)
}
if ( !file.exists(file_name_test) ) {
  dir.create("data/original")
  download.file(file_url_test, file_name_test)
}

```

Initial exploration reveals some missing datapoints and error messages from micrsoft  excel spreadsheet calculations.  Those are all converted to 'NA'.  Then any column containing NA is removed from the calculation.  AFter that, all non-numeric columns are removed as well as columns used for labelling i.e. "window_num". Finaly, using caret package, all correlated columns are removed.  The nearZero function was also used but did not reduce the column count so it was commented out.  160 colunms was reduced to 46 including the classe column.

```{r clean_data_basic,message=FALSE, warning=FALSE, cache=TRUE}

library(data.table)

# Read the dataset using `fread()` in `data.table` package.
train_raw <- fread(file_name_train, na.strings = c("NA","#DIV/0!",""))
train_1 <-train_raw 
train_raw <- train_raw[,colSums(is.na(train_raw)) == 0,with=F]
train_2 <- train_raw
train_raw <- train_raw[,sapply(train_raw, is.numeric),with=F]
train_3 <- train_raw

# Even with caret, we failed to remove timestamp and window label
indentifiers <- c("V1", "user_name","raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp","new_window","num_window")
train_raw <- train_raw[,!colnames(train_raw) %in% indentifiers,with=F] 

```


``` {r clean_data_advanced, message=FALSE, warning=FALSE, cache=TRUE}
library(caret)

filterColumnsPart1 <- function(x) {
  temp <- nearZeroVar(x, saveMetrics=T)
  #remove columns with low variance, a.k.a. statistically insignificant
  x <- x[,!temp$nzv & !temp$zeroVar,with=F]
}
filterColumnsPart2 <- function(x){
  #remove correlated columns
  cm <- cor(x, use="pairwise.complete.obs")
  #print(summary(cm))
  temp <- findCorrelation(cm)
  #print(dim(temp))
  #print(head(temp))
  x <- x[,-temp,with=F]
}
filterColumnsPart3 <- function(x){
  temp <-preProcess(x, method="pca")
  x <- predict(temp,x)
}
train_raw <-filterColumnsPart2(train_raw)
train_6 <- train_raw
train_raw$classe <- as.factor(train_1$classe)

#dim(train_raw)
cc  <- colnames(train_raw)
c <- colnames(train_3)
#print(setdiff(cc,c))
#dim(train_1)
#dim(train_raw)
```

```{r submission_init}
library(data.table)
test_raw <- fread(file_name_test, na.strings = c("NA","#DIV/0!",""))
#test_raw <-test_raw[,colnames(train_raw)]
```

Creating a training set and testing set from  the data

```{r pred_data, message=FALSE, warning=FALSE, cache=TRUE}
library(randomForest)

inTrain <- createDataPartition(train_raw$classe, p=0.7, list=FALSE)
training <- train_raw[inTrain,]
testing <- train_raw[-inTrain,]
#dim(training)
#dim(testing)
#summary(train_raw)

```
``` {r message=FALSE, warning=FALSE, cache=TRUE}
test_2<-function(training,testing,train_method) {
  temp <- train(classe ~ ., 
              data=training, 
              method=train_method, 
              preProcess="pca", 
              metric="Accuracy",
              allowParallel=T,
              trControl = trainControl(preProcOptions = list(thresh = 0.8)))
  finalModel<-temp$finalModel
  trainAccuracy <- temp$results$Accuracy[1]
  print(train_method)
  print(finalModel)
  print(train_method)
  print(trainAccuracy)
  testing$classe2 <- predict(temp, newdata=testing)
  cm <-confusionMatrix(data=testing$classe2, testing$classe)
  print(cm)
#  testSet$predict <- predict(rf, newdata=testSet )
 temp
}
set.seed(1235)
#method_rpart <- test_2(training,testing,"rpart")
#method_glm   <- test_2(training,testing,"glm")
method_rf    <- test_2(training,testing,"rf")

```

Finally part two of the submission

```{r submission_process, message=FALSE, warning=FALSE}

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("data/submission/",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
submissions <- predict(method_rf, test_raw)
print(submissions)
dir.create("data/submission")

pml_write_files(submissions)
```

###Appendix A : Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. This is the "classe" variable in the training set. 

###Appendix B : Data Source

To evaluate the HTML file you may have to download the repo and open the compiled HTML document. 

Alternatively if they have submitted a repo with a gh-pages branch, you may be able to view the HTML page on the web. If the repo is: 

https://github.com/DataScienceSpecialization/courses/tree/master/08_PracticalMachineLearning/001predictionMotivation

then you can view the HTML page here: 

http://datasciencespecialization.github.io/courses/08_PracticalMachineLearning/001predictionMotivation/

###Appendix C : What you should submit

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 

###Appendix D : Reproducibility 

Due to security concerns with the exchange of R code, your code will not be run during the evaluation by your classmates. Please be sure that if they download the repo, they will be able to view the compiled HTML version of your analysis. 

- Has the student submitted a github repo?

Does the submission build a machine learning algorithm to predict activity quality from activity monitors?


- Do the authors describe what they expect the out of sample error to be and estimate the error appropriately with cross-validation?

